Neural Connectome of the Ctenophore Statocyst

### 2025-07-11 manuscript/Manuscript.qmd
Ms: Video 7 add reference in Results

For vEM analysis, we used a five-day-old cydippid larva of *Mnemiopsis leidyi*. 
To ensure optimal ultrastructural preservation, the specimen was prepared by high-pressure freezing, followed by freeze substitution and Epon embedding. 
From the aboral tip, we collected \~1,000 ultra-thin (50 nm) serial sections as ribbons, targeting only the region containing the aboral organ. 
These were imaged using a Zeiss Gemini 500 SEM at a pixel size (xy) of 2.0 nm. 
We imaged only the region containing the aboral organ. Subsequently, images from 620 layers were stitched and aligned in TrakEM2 [@cardona2012]. 
The final vEM dataset encompassed a volume of 60 μm × 40 μm × 30 μm. 
We skeletonised and annotated all cells in this dataset in CATMAID [@Saalfeld2009; @Schneider-Mizell2016]. 
We reconstructed 972 cells with intact nuclei and somata. 
Among these, 898 were located within the aboral organ (AO), including balancers and the aboral nerve net neurons. 
Due to the compact and mostly self-contained organization of the AO, these cells were largely contained within the imaged volume.
We classified these cells into cell types (Figure 1--figure supplement 1, Figure 1--figure supplement 2).

@@

To investigate whether balancer cilia are coordinated by neural inputs, we did high-speed video microscopy recordings to analyze their activity patterns across body axes. 
Previous studies [@Tamm1980; @Tamm1982; @Lowe1997] have established that balancer cilia function as mechanoreceptors, with their beating frequency modulated by inclination. 
Tamm also suggested that differences in statolith morphology and the shape of balancer cilia between the tentacular and sagittal planes could lead to different forces exerted on cilia by the statolith [@Tamm_2014c; @tamm2015]. 
To further explore this, we used a tilted microscope with a vertical stage where we mounted immobilised cydippid larvae with their aboral-oral axis aligned in different orientations relative to the gravity vector. 
In all cases, only larvae whose aboral–oral axis was approximately parallel to the gravity vector (mouth facing downward within a 0–20° deviation) were included for analysis. 
This ensured that all balancer cilia experienced comparable gravitational input from the statolith, allowing for an unbiased comparison across quadrants.
We then compared larvae oriented in different angles and either with their sagittal or tentacular plane parallel to the sample stage (Figure 4A).

@@

This orientation corresponds to a “balancer-up” posture, in which the aboral organ faces upward. We selected this configuration not only to ensure consistent stimulation from the statolith across individuals, but also because it facilitates intuitive interpretation of balancer function—as if balancing a stone on top. Moreover, this posture matches the orientation used in electron microscopy and was technically advantageous for high-speed imaging under vertical-stage alignment. Using a consistent orientation throughout minimized potential confusion when comparing ultrastructural and functional data.

Because tilt angle strongly affects the beat frequency of balancer cilia via asymmetric mechanical loading by the statolith [@Tamm1980], our quantitative analysis was restricted to recordings that met the above orientation criterion (tilt < 20°). In this orientation, gravitational input from the statolith is approximately symmetrical across the left and right balancers, allowing meaningful comparisons of their activity. To illustrate how excessive tilt induces asymmetric stimulation, Video 7 presents two examples of larvae outside the permitted range: one tilted by ~135° in the sagittal plane and the other by ~60° in the tentacular plane. In both cases, strong asymmetries in balancer ciliary beating are evident, consistent with previous observations of gravity-induced geotactic responses [@Tamm1980]. To illustrate how excessive tilt induces asymmetric stimulation, these examples demonstrate why we restricted our analysis to near-vertical orientations, where gravitational input is approximately balanced across the balancers.

>>>>

For vEM analysis, we used a five-day-old cydippid larva of *Mnemiopsis leidyi*. 
To ensure optimal ultrastructural preservation, the specimen was prepared by high-pressure freezing, followed by freeze substitution and Epon embedding. 
From the aboral tip, we collected \~1,000 ultra-thin (50 nm) serial sections as ribbons, targeting only the region containing the aboral organ. 
These were imaged using a Zeiss Gemini 500 SEM at a pixel size (xy) of 2.0 nm. 
We imaged only the region containing the aboral organ. Subsequently, images from 620 layers were stitched and aligned in TrakEM2 [@cardona2012]. 
The final vEM dataset encompassed a volume of 60 μm × 40 μm × 30 μm. 
We skeletonised and annotated all cells in this dataset in CATMAID [@Saalfeld2009; @Schneider-Mizell2016]. 
We reconstructed 972 cells with intact nuclei and somata. 
Among these, 898 were located within the aboral organ (AO), including balancers and the aboral nerve net neurons (Video 1). 
Due to the compact and mostly self-contained organization of the AO, these cells were largely contained within the imaged volume.
We classified these cells into cell types (Figure 1--figure supplement 1, Figure 1--figure supplement 2).

@@

To investigate whether balancer cilia are coordinated by neural inputs, we did high-speed video microscopy recordings to analyze their activity patterns across body axes. 
Previous studies [@Tamm1980; @Tamm1982; @Lowe1997] have established that balancer cilia function as mechanoreceptors, with their beating frequency modulated by inclination.  
Tamm also suggested that differences in statolith morphology and the shape of balancer cilia between the tentacular and sagittal planes could lead to different forces exerted on cilia by the statolith [@Tamm_2014c; @tamm2015]. 
To further explore this, we used a tilted microscope with a vertical stage where we mounted immobilised cydippid larvae with their aboral-oral axis aligned in different orientations relative to the gravity vector (Videos 3-7). We could confirm earlier finding that the beating frequency of balancer cilia is dependent on the inclination (Video 7). Our aim here was to focus on spontaneous ciliary beating under conditions of equal load from the statolith across quadrants.
Therefore, we only analysed larvae whose aboral–oral axis was approximately parallel to the gravity vector (mouth facing downward within a 0–20° deviation). 
We then compared larvae oriented either with their sagittal or tentacular plane parallel to the sample stage (Figure 4A).

@@

This orientation corresponds to a “balancer-up” orientation, in which the aboral organ faces upward. We selected this configuration not only to ensure consistent stimulation from the statolith across individuals, but also because it facilitates intuitive interpretation of balancer function—as if balancing a stone on top. Moreover, this orientation matches the orientation used in electron microscopy and was technically advantageous for high-speed imaging under vertical-stage alignment. Using a consistent orientation throughout minimized potential confusion when comparing ultrastructural and functional data.

Because the tilt angle strongly affects the beat frequency of balancer cilia via asymmetric mechanical loading by the statolith [@Tamm1980], our quantitative analysis was restricted to recordings that met the above orientation criterion (tilt < 20°). In this orientation, gravitational input from the statolith is approximately symmetrical across the left and right balancers, allowing meaningful comparisons of their activity. To illustrate how excessive tilt induces asymmetric stimulation, Video 7 presents two examples of larvae outside the permitted range: one tilted by ~135° in the sagittal plane and the other by ~60° in the tentacular plane. In both cases, strong asymmetries in balancer ciliary beating are evident, consistent with previous observations of gravity-induced geotactic responses [@Tamm1980].
